---
title: "The 'jack' package: Jack polynomials"
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check-valgrind](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
juliaeval <- TRUE
```

```{r}
library(jack)
library(microbenchmark)
```

Schur polynomials have applications in combinatorics and zonal polynomials have
applications in multivariate statistics. They are particular cases of
[Jack polynomials](https://en.wikipedia.org/wiki/Jack_function). This package
allows to evaluate these polynomials. It can also compute their symbolic form.

The  functions `JackPol`, `ZonalPol`, `ZonalQPol` and `SchurPol` respectively 
return the Jack polynomial, the zonal polynomial, the quaternionic zonal 
polynomial, and the Schur polynomial. 

Each of these polynomials corresponds is given by a positive integer, the 
number of variables, and an integer partition, the `lambda` argument; the 
Jack polynomial has one more parameter, the `alpha` argument, a 
positive number.

To get an exact symbolic polynomial with `JackPol`, you have to supply a 
`bigq` rational number for the parameter `alpha`:

```{r}
jpol <- JackPol(2, lambda = c(3, 1), alpha = gmp::as.bigq("2/5"))
jpol
```

This is a `qspray` object, from the [**qspray**](https://github.com/stla/qspray) 
package. Here is how you can evaluate this polynomial:

```{r}
qspray::evalQspray(jpol, c("2", "3/2"))
```

By default, `ZonalPol`, `ZonalQPol` and `SchurPol` return exact symbolic 
polynomials.

```{r}
zpol <- ZonalPol(2, lambda = c(3, 1))
zpol
```

It is also possible to convert a `qspray` polynomial to a function whose 
evaluation is performed by the **Ryacas** package:

```{r}
zyacas <- as.function(zpol)
```

You can provide the values of the variables of this function as numbers or 
character strings:

```{r}
zyacas(2, "3/2")
```

You can even pass a variable name to this function:

```{r}
zyacas("x", "x")
```

If you want to substitute a variable with a complex number, use a character 
string which represents this number, with `I` denoting the imaginary unit:

```{r}
zyacas("2 + 2*I", "2/3")
```


## Jack polynomials with Julia

As of version 2.0.0, the Jack polynomials can be calculated with Julia. The 
speed is amazing:

```{r, eval=juliaeval}
julia <- Jack_julia()
x <- c(1/2, 2/3, 1, 2/3, -1, -2, 1)
lambda <- c(5, 3, 2, 2, 1)
alpha <- 3
print(
  microbenchmark(
        R = Jack(x, lambda, alpha),
    Julia = julia$Jack(x, lambda, alpha),
    times = 6L,
    unit  = "seconds"
  ),
  signif = 6L
)
```

`Jack_julia()` returns a list of functions. `ZonalPol`, `ZonalQPol` and 
`SchurPol` always return an exact expression of the polynomial, i.e. with 
rational coefficients (integers for `SchurPol`). If you want an exact 
expression with `JackPol`, you have to give a rational number for the argument 
`alpha`, as a character string:

```{r, eval=juliaeval}
JP <- julia$JackPol(m = 2, lambda = c(3, 1), alpha = "2/5")
JP
```

Again, Julia is faster:

```{r, eval=juliaeval}
n <- 5
lambda <- c(4, 3, 3)
alpha <- "2/3"
alphaq <- gmp::as.bigq(alpha)
print(
  microbenchmark(
        R = JackPol(n, lambda, alphaq),
    Julia = julia$JackPol(n, lambda, alpha),
    times = 6L
  ),
signif = 2L)
```


## 'Rcpp' implementation of the polynomials

As of version 5.0.0, a 'Rcpp' implementation of the polynomials is provided by 
the package. It is faster than Julia (though I didn't compare in pure Julia - 
the Julia execution time is slowed down by the 'JuliaConnectoR' package):

```{r, eval=juliaeval}
n <- 5
lambda <- c(4, 3, 3, 2)
print(
  microbenchmark(
     Rcpp = SchurPolCPP(n, lambda),
    Julia = julia$SchurPol(n, lambda),
    times = 6L
  ), 
signif = 2L)
```

```{r, eval=juliaeval}
n <- 5
lambda <- c(4, 3, 3, 2)
alpha <- "2/3"
print(
  microbenchmark(
     Rcpp = JackPolCPP(n, lambda, alpha),
    Julia = julia$JackPol(n, lambda, alpha),
    times = 6L
  ), 
signif = 2L)
```

As of version 5.1.0, there's also a 'Rcpp' implementation of the evaluation of 
the polynomials.

```{r, eval=juliaeval}
x <- c("1/2", "2/3", "1", "2/3", "1", "5/4")
lambda <- c(5, 3, 2, 2, 1)
alpha <- "3"
print(
  microbenchmark(
        R = Jack(gmp::as.bigq(x), lambda, gmp::as.bigq(alpha)),
     Rcpp = JackCPP(x, lambda, alpha),    
    times = 10L,
    unit  = "seconds"
  ),
  signif = 6L
)
```

```{r, eval=juliaeval}
JuliaConnectoR::stopJulia()
```
