---
title: "The 'jack' package: Jack polynomials"
output: github_document
editor_options: 
  chunk_output_type: console
---

***Jack, zonal, Schur, and other symmetric polynomials.***

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check-valgrind](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r, message=FALSE}
library(jack)
```

*Schur polynomials* have applications in combinatorics and *zonal polynomials* 
have applications in multivariate statistics. They are particular cases of
[*Jack polynomials*][JackPolynomials]. The original purpose of this package was 
the evaluation and the computation in symbolic form of these polynomials. 
Now it contains much more stuff dealing with multivariate symmetric polynomials.


## Breaking change in version 6.0.0

In version 6.0.0, each function whose name ended with the suffix `CPP` 
(`JackCPP`, `JackPolCPP`, etc.) has been renamed by removing this suffix, 
and the functions `Jack`, `JackPol`, etc. have been renamed by adding the 
suffix `R` to their name: `JackR`, `JackPolR`, etc. The reason of these 
changes is that a name like `Jack` is more appealing than `JackCPP` and 
it is more sensible to assign the more appealing names to the functions 
implemented with **Rcpp** since they are highly more efficient. The interest 
of the functions `JackR`, `JackPolR`, etc. is meager.


## Getting the polynomials

The functions `JackPol`, `ZonalPol`, `ZonalQPol` and `SchurPol` respectively 
return the Jack polynomial, the zonal polynomial, the quaternionic zonal 
polynomial, and the Schur polynomial. 

Each of these polynomials is given by a positive integer, the 
number of variables (the `n` argument), and an integer partition (the 
`lambda` argument); the Jack polynomial has a parameter in addition, the 
`alpha` argument, a number called the *Jack parameter*.

Actually there are four possible Jack polynomials for a given Jack parameter 
and a given integer partition: the $J$-polynomial, the $P$-polynomial, the 
$Q$-polynomial and the $C$-polynomial. You can specify which one you want 
with the `which` argument, which is set to `"J"` by default. These four 
polynomials differ only by a constant factor.

To get a Jack polynomial with `JackPol`, you have to supply the Jack parameter
as a `bigq` rational number or as a character string representing a fraction, 
e.g. `"2/5"`:

```{r}
jpol <- JackPol(2, lambda = c(3, 1), alpha = "2/5")
jpol
```

This is a `qspray` object, from the [**qspray** package][qspray]. Here is how 
you can evaluate this polynomial:

```{r}
evalQspray(jpol, c("2", "3/2"))
```

It is also possible to convert a `qspray` polynomial to a function whose 
evaluation is performed by the **Ryacas** package:

```{r}
jyacas <- as.function(jpol)
```

You can provide the values of the variables of this function as numbers or 
character strings:

```{r}
jyacas(2, "3/2")
```

You can even pass a variable name to this function:

```{r}
jyacas("x", "x")
```

If you want to substitute a complex number to a variable, use a character 
string which represents this number, with `I` denoting the imaginary unit:

```{r}
jyacas("2 + 2*I", "2/3 + I/4")
```

It is also possible to evaluate a `qspray` polynomial for some complex values 
of the variables with `evalQspray`. You have to separate the real parts and the
imaginary parts:

```{r}
evalQspray(jpol, values_re = c("2", "2/3"), values_im = c("2", "1/4"))
```


## Direct evaluation of the polynomials

If you just have to evaluate a Jack polynomial, you don't need to resort to 
a `qspray` polynomial: you can use the functions `Jack`, `Zonal`, `ZonalQ` or
`Schur`, which directly evaluate the polynomial; this is much more efficient 
than computing the `qspray` polynomial and then applying `evalQspray`.

```{r}
Jack(c("2", "3/2"), lambda = c(3, 1), alpha = "2/5")
```

However, if you have to evaluate a Jack polynomial for several values, it 
could be better to resort to the `qspray` polynomial.


## Skew Schur polynomials

As of version 6.0.0, the package is able to compute the skew Schur polynomials, 
with the function `SkewSchurPol`.


## Symbolic Jack parameter

As of version 6.0.0, it is possible to get a Jack polynomial with a symbolic 
Jack parameter in its coefficients, thanks to the 
[**symbolicQspray** package][symbolicQspray]. 

```{r}
( J <- JackSymPol(2, lambda = c(3, 1)) )
```

This is a `symbolicQspray` object, from the **symbolicQspray** package.

A `symbolicQspray` object corresponds to a multivariate polynomial whose 
coefficients are fractions of polynomials with rational coefficients. The 
variables of these fractions of polynomials can be seen as some parameters.
The Jack polynomials fit into this category: from their definition, their 
coefficients are fractions of polynomials in the Jack parameter. However you 
can see in the above output that for this example, the coefficients are 
*polynomials* in the Jack parameter (`a`): there's no fraction. Actually this 
fact is always true for the Jack $J$-polynomials. This is an established fact 
and it is not obvious (it is a consequence of the 
[Knop & Sahi formula][KnopSahi]).

You can substitute a value to the Jack parameter with the help of the 
`substituteParameters` function:

```{r}
( J5 <- substituteParameters(J, 5) )
J5 == JackPol(2, lambda = c(3, 1), alpha = "5")
```

Note that you can change the letters used to denote the variables. By default, 
the Jack parameter is denoted by `a` and the variables are denoted by `X`, `Y`, 
`Z` if there are no more than three variables, otherwise they are denoted by 
`X1`, `X2`, ... Here is how to change these symbols:

```{r}
showSymbolicQsprayOption(J, "a") <- "alpha"
showSymbolicQsprayOption(J, "X") <- "x"
J
```

If you want to have the variables denoted by `x` and `y`, do:

```{r}
showSymbolicQsprayOption(J, "showMonomial") <- showMonomialXYZ(c("x", "y"))
J
```


## Compact expression of Jack polynomials

The expression of a Jack polynomial in the canonical basis can be long. Since 
these polynomials are symmetric, one can get a considerably shorter expression 
by writing the polynomial as a linear combination of the monomial symmetric 
polynomials. This is what the function `compactSymmetricQspray` does:

```{r}
( J <- JackPol(3, lambda = c(4, 3, 1), alpha = "2") )
compactSymmetricQspray(J) |> cat()
```

The function `compactSymmetricQspray` is also applicable to a `symbolicQspray`
object, like a Jack polynomial with symbolic Jack parameter.

It is easy to figure out what is a monomial symmetric polynomial: `M[i, j, k]` 
is the sum of all monomials `x^p.y^q.z^r` where `(p, q, r)` is a permutation 
of `(i, j, k)`.

The "compact expression" of a Jack polynomial with `n` variables does not 
depend on `n` if `n >= sum(lambda)`:
```{r}
lambda <- c(3, 1)
alpha <- "3"
J4 <- JackPol(4, lambda, alpha)
J9 <- JackPol(9, lambda, alpha)
compactSymmetricQspray(J4) |> cat()
compactSymmetricQspray(J9) |> cat()
```

In fact I'm not sure the Jack polynomial makes sense when `n < sum(lambda)`.


## Hall inner product

The **qspray** package provides a function to compute the Hall inner product 
of two symmetric polynomials, namely `HallInnerProduct`. This is the 
generalized Hall inner product, the one with a parameter $\alpha$. It is known 
that the Jack polynomials with parameter $\alpha$ are orthogonal for the Hall 
inner product with parameter $\alpha$. Let's give a try:

```{r}
alpha <- "3"
J1 <- JackPol(4, lambda = c(3, 1), alpha, which = "P")
J2 <- JackPol(4, lambda = c(2, 2), alpha, which = "P")
HallInnerProduct(J1, J2, alpha)
HallInnerProduct(J1, J1, alpha)
HallInnerProduct(J2, J2, alpha)
```
If you set `alpha=NULL` in `HallInnerProduct`, you get the Hall inner product 
with symbolic parameter $\alpha$:
```{r}
HallInnerProduct(J1, J1, alpha = NULL)
```

This is a `qspray` object. The Hall inner product is always polynomial in 
$\alpha$.

It is also possible to get the Hall inner product of two `symbolicQspray` 
polynomials. Take for example a Jack polynomial with symbolic parameter:
```{r}
J <- JackSymPol(4, lambda = c(3, 1), which = "P")
showSymbolicQsprayOption(J, "a") <- "t"
HallInnerProduct(J, J, alpha = 2)
```
We use `t` to display the Jack parameter and not `alpha` so that there is no 
confusion between the Jack parameter and the parameter of the Hall product.

Now, what happens if we compute the symbolic Hall inner product of this Jack
polynomial with itself, that is, if we run 
`HallInnerProduct(J, J, alpha = NULL)`? Let's see:
```{r}
( Hip <- HallInnerProduct(J, J, alpha = NULL) )
```

We get the Hall inner product of the Jack polynomial with itself, with two 
symbolic parameters: the Jack parameter displayed as `t` and the parameter 
of the Hall product displayed as `alpha`. This is a `symbolicQspray` object.

Now one could be interested in the symbolic Hall inner product of the Jack 
polynomial with itself for the case when the Jack parameter and the parameter 
of the Hall product coincide, that is, to set `alpha=t` in the `symbolicQspray` 
polynomial that we named `Hip`. One can get it as follows:
```{r}
changeVariables(Hip, list(qlone(1)))
```

This is rather a trick. The `changeVariables` function allows to replace the 
variables of a `symbolicQspray` polynomial with the new variables given as a 
list in its second argument. The `Hip` polynomial has only one variable, 
`alpha`, and it has one parameter, `t`. This parameter `t` is the polynomial 
variable of the `ratioOfQsprays` coefficients of `Hip`. Technically this is 
a `qspray` object: this is `qlone(1)`. So we provided `list(qlone(1))` as the 
list of new variables. This corresponds to set `alpha=t`. The usage of the 
`changeVariables` is a bit deflected, because `qlone(1)` is not a new variable 
for `Hip`, this is a constant.


## Laplace-Beltrami operator

Just to illustrate the possibilities of the packages involved in the **jack** 
package (**qspray**, **ratioOfQsprays**, **symbolicQspray**), let us check 
that the Jack polynomials are eigenpolynomials for the 
[Laplace-Beltrami operator][LaplaceBeltrami] on the space of homogeneous 
symmetric polynomials.

```{r}
LaplaceBeltrami <- function(qspray, alpha) {
  n <- numberOfVariables(qspray)
  derivatives1 <- lapply(seq_len(n), function(i) {
    derivQspray(qspray, i)
  })
  derivatives2 <- lapply(seq_len(n), function(i) {
    derivQspray(derivatives1[[i]], i)
  })
  x <- lapply(seq_len(n), qlone) # x_1, x_2, ..., x_n
  # first term
  out1 <- 0L
  for(i in seq_len(n)) {
    out1 <- out1 + alpha * x[[i]]^2 * derivatives2[[i]]
  }
  # second term
  out2 <- 0L
  for(i in seq_len(n)) {
    for(j in seq_len(n)) {
      if(i != j) {
        out2 <- out2 + x[[i]]^2 * derivatives1[[i]] / (x[[i]] - x[[j]])
      }
    }
  }
  # at this step, `out2` is a `ratioOfQsprays` object, because of the divisions
  # by `x[[i]] - x[[j]]`; but actually its denominator is 1 because of some
  # simplifications and then we extract its numerator to get a `qspray` object
  out2 <- getNumerator(out2)
  out1/2 + out2
}
```

```{r}
alpha <- "3"
J <- JackPol(4, c(2, 2), alpha)
collinearQsprays(
  qspray1 = LaplaceBeltrami(J, alpha), 
  qspray2 = J
)
```


## Other symmetric polynomials

Many other symmetric multivariate polynomials have been introduced in version 6.1.0. 
Let's see a couple of them.


### Skew Jack polynomials

The skew Jack polynomials are now available. They generalize the skew Schur 
polynomials. The skew Schur polynomial associated to some skew partition is 
the skew Jack $P$-polynomial with Jack parameter $\alpha=1$ associated to the 
same skew partition:

```{r}
n <- 3
lambda <- c(3, 3)
mu <- c(2, 1)
skewSchurPoly <- SkewSchurPol(n, lambda, mu)
skewJackPoly <- SkewJackPol(n, lambda, mu, alpha = 1, which = "P")
skewSchurPoly == skewJackPoly
```


### $t$-Schur polynomials

The $t$-Schur polynomials depend on a single parameter usually denoted by $t$ and 
their coefficients are polynomials in this parameter. They yield the Schur polynomials
when substituting $t$ with $0$:

```{r}
n <- 3
lambda <- c(2, 2)
tSchurPoly <- tSchurPol(n, lambda)
substituteParameters(tSchurPoly, values = 0) == SchurPol(n, lambda)
```


### Hall-Littlewood polynomials

Similarly to the $t$-Schur polynomials, the Hall-Littlewood polynomials depend on a
single parameter usually denoted by $t$ and their coefficients are polynomials in 
this parameter. The Hall-Littlewood $P$-polynomials yield the Schur polynomials
when substituting $t$ with $0$:

```{r}
n <- 3
lambda <- c(2, 2)
hlPoly <- HallLittlewoodPol(n, lambda, which = "P")
substituteParameters(hlPoly, values = 0) == SchurPol(n, lambda)
```


### Macdonald polynomials

The Macdonald polynomials depend on two parameters usually denoted by $q$ and $t$.
Their coefficients are not polynomials in $q$ and $t$ in general, they are ratios 
of polynomials in $q$ and $t$. They yield the Hall-Littlewood polynomials when 
substituting $q$ with $0$:

```{r}
n <- 3
lambda <- c(2, 2)
macPoly <- MacdonaldPol(n, lambda)
hlPoly <- HallLittlewoodPol(n, lambda)
changeParameters(macPoly, list(0, qlone(1))) == hlPoly
```




<!-- -------------------- links -------------------- -->

[JackPolynomials]: https://en.wikipedia.org/wiki/Jack_function "Jack polynomials on Wikipedia"

[qspray]: https://github.com/stla/qspray "the 'qspray' package on Github"

[symbolicQspray]: https://github.com/stla/symbolicQspray "the 'symbolicQspray' package on Github"

[KnopSahi]: https://en.wikipedia.org/wiki/Jack_function#Combinatorial_formula "Knop & Sahi formula"

[LaplaceBeltrami]: https://math.mit.edu/~rstan/pubs/pubfiles/73.pdf "Laplace-Beltrami operator"
