---
title: "The 'jack' package: Jack polynomials"
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check-valgrind](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r, message=FALSE}
library(jack)
library(microbenchmark)
```

Schur polynomials have applications in combinatorics and zonal polynomials have
applications in multivariate statistics. They are particular cases of
[Jack polynomials](https://en.wikipedia.org/wiki/Jack_function). This package
allows to evaluate these polynomials. It can also compute their symbolic form.

The  functions `JackPol`, `ZonalPol`, `ZonalQPol` and `SchurPol` respectively 
return the Jack polynomial, the zonal polynomial, the quaternionic zonal 
polynomial, and the Schur polynomial. 

Each of these polynomials is given by a positive integer, the 
number of variables, and an integer partition, the `lambda` argument; the 
Jack polynomial has one more parameter, the `alpha` argument, a number called 
the *Jack parameter*.

To get an exact symbolic polynomial with `JackPol`, you have to supply a 
`bigq` rational number for the Jack parameter `alpha`:

```{r}
jpol <- JackPol(2, lambda = c(3, 1), alpha = gmp::as.bigq("2/5"))
jpol
```

This is a `qspray` object, from the [**qspray**](https://github.com/stla/qspray) 
package. Here is how you can evaluate this polynomial:

```{r}
evalQspray(jpol, c("2", "3/2"))
```

By default, `ZonalPol`, `ZonalQPol` and `SchurPol` return exact symbolic 
polynomials.

```{r}
zpol <- ZonalPol(2, lambda = c(3, 1))
zpol
```

It is also possible to convert a `qspray` polynomial to a function whose 
evaluation is performed by the **Ryacas** package:

```{r}
zyacas <- as.function(zpol)
```

You can provide the values of the variables of this function as numbers or 
character strings:

```{r}
zyacas(2, "3/2")
```

You can even pass a variable name to this function:

```{r}
zyacas("x", "x")
```

If you want to substitute a variable with a complex number, use a character 
string which represents this number, with `I` denoting the imaginary unit:

```{r}
zyacas("2 + 2*I", "2/3")
```


## Jack polynomials with Julia

As of version 2.0.0, it was possible to calculate the Jack polynomials with 
Julia. This feature has been removed in version 5.3.0. Use the Julia package
**JackPolynomials.jl** instead.


## 'Rcpp' implementation of the polynomials

As of version 5.0.0, a 'Rcpp' implementation of the polynomials is provided by 
the package.

As of version 5.1.0, there's also a 'Rcpp' implementation of the evaluation of 
the polynomials.

```{r, eval=FALSE}
x <- c("1/2", "2/3", "1", "2/3", "1", "5/4")
lambda <- c(5, 3, 2, 2, 1)
alpha <- "3"
print(
  microbenchmark(
        R = Jack(gmp::as.bigq(x), lambda, gmp::as.bigq(alpha)),
     Rcpp = JackCPP(x, lambda, alpha),
    times = 5L,
    unit  = "seconds"
  ),
  signif = 2L
)
```


## Skew Schur polynomials

As of version 6.0.0, the package is able to compute the skew Schur polynomials, 
with the function `SkewSchurPol`.


## Symbolic Jack parameter

As of version 6.0.0, it is possible to get a Jack polynomial with a symbolic 
Jack parameter in its coefficients, with the `JackSymPol` function:

```{r}
( J <- JackSymPol(2, lambda = c(3, 1)) )
```

This is a `symbolicQspray` object, from the
[**symbolicQspray**](https://github.com/stla/symbolicQspray) package.

A `symbolicQspray` object corresponds to a multivariate polynomial whose 
coefficients are fractions of polynomials with rational coefficients. 
The Jack polynomials fit into this category: from their definition, their 
coefficients are fractions of polynomials in the Jack parameter. However you 
can see in the above output that for this example, the coefficients are 
*polynomials* in the Jack parameter (`a`): there's no fraction. Actually this is 
always true for any Jack polynomial. This fact is established and it is not 
obvious.

Note that you can change the letters used to denote the variables. By default, 
the Jack parameter is denoted by `a` and the variables are denoted by `X1`, 
`X2`, ... Here is how to change these symbols:

```{r}
showSymbolicQsprayOption(J, "a") <- "alpha"
showSymbolicQsprayOption(J, "X") <- "x"
J
```


## Compact expression of Jack polynomials

The expression of a Jack polynomial in the canonical basis can be long. Since 
these polynomials are symmetric, one can get a considerably shorter expression 
by writing the polynomial as a linear combination of the monomial symmetric 
polynomials. This is what the function `compactSymmetricQspray` does:

```{r}
( J <- JackPolCPP(3, lambda = c(4, 3, 1), alpha = "2") )
cat(compactSymmetricQspray(J))
```

The function `compactSymmetricQspray` is from the **qspray** package but it is 
also possible to apply it to a `symbolicQspray` object, like a Jack polynomial 
with symbolic Jack parameter.


