---
title: "The 'jack' package: Jack polynomials"
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check.yaml)
[![R-CMD-check-valgrind](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml/badge.svg)](https://github.com/stla/jackR/actions/workflows/R-CMD-check-valgrind.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE)
```

```{r, message=FALSE}
library(jack)
```

Schur polynomials have applications in combinatorics and zonal polynomials have
applications in multivariate statistics. They are particular cases of
[Jack polynomials][JackPolynomials]. This package allows to evaluate these
polynomials and also to compute them in symbolic form.


## Breaking change in version 6.0.0

In version 6.0.0, each function whose name ended with the suffix `CPP` 
(`JackCPP`, `JackPolCPP`, etc.) has been renamed by removing this suffix, 
and the functions `Jack`, `JackPol`, etc. have been renamed by adding the 
suffix `R` to their name: `JackR`, `JackPolR`, etc. The reason of these 
changes is that a name like `Jack` is more appealing than `JackCPP` and 
it is more sensible to assign the more appealing names to the functions 
implemented with **Rcpp** since they are highly more efficient. The interest 
of the functions `JackR`, `JackPolR`, etc. is meager.


## Getting the polynomials

The functions `JackPol`, `ZonalPol`, `ZonalQPol` and `SchurPol` respectively 
return the Jack polynomial, the zonal polynomial, the quaternionic zonal 
polynomial, and the Schur polynomial. 

Each of these polynomials is given by a positive integer, the 
number of variables (the `n` argument), and an integer partition (the 
`lambda` argument); the Jack polynomial has a parameter in addition, the 
`alpha` argument, a number called the *Jack parameter*.

To get a Jack polynomial with `JackPol`, you have to supply the Jack parameter
as a `bigq` rational number or as a character string representing a fraction, 
e.g. `"2/5"`:

```{r}
jpol <- JackPol(2, lambda = c(3, 1), alpha = "2/5")
jpol
```

This is a `qspray` object, from the [**qspray** package][qspray]. Here is how 
you can evaluate this polynomial:

```{r}
evalQspray(jpol, c("2", "3/2"))
```

It is also possible to convert a `qspray` polynomial to a function whose 
evaluation is performed by the **Ryacas** package:

```{r}
jyacas <- as.function(jpol)
```

You can provide the values of the variables of this function as numbers or 
character strings:

```{r}
jyacas(2, "3/2")
```

You can even pass a variable name to this function:

```{r}
jyacas("x", "x")
```

If you want to substitute a variable with a complex number, use a character 
string which represents this number, with `I` denoting the imaginary unit:

```{r}
jyacas("2 + 2*I", "2/3")
```


## Direct evaluation of the polynomials

If you just have to evaluate a Jack polynomial, you don't need to resort to 
a `qspray` polynomial: you can use the functions `Jack`, `Zonal`, `ZonalQ` or
`Schur`, which directly evaluate the polynomial; this is much more efficient 
than computing the `qspray` polynomial and then applying `evalQspray`.

```{r}
Jack(c("2", "3/2"), lambda = c(3, 1), alpha = "2/5")
```

However, if you have to evaluate a Jack polynomial for several values, it 
could be better to resort to the `qspray` polynomial.


## Skew Schur polynomials

As of version 6.0.0, the package is able to compute the skew Schur polynomials, 
with the function `SkewSchurPol`.


## Symbolic Jack parameter

As of version 6.0.0, it is possible to get a Jack polynomial with a symbolic 
Jack parameter in its coefficients, thanks to the 
[**symbolicQspray** package][symbolicQspray]. 

```{r}
( J <- JackSymPol(2, lambda = c(3, 1)) )
```

This is a `symbolicQspray` object, from the **symbolicQspray** package.

A `symbolicQspray` object corresponds to a multivariate polynomial whose 
coefficients are fractions of polynomials with rational coefficients. The 
variables of these fractions of polynomials can be seen as some parameters.
The Jack polynomials fit into this category: from their definition, their 
coefficients are fractions of polynomials in the Jack parameter. However you 
can see in the above output that for this example, the coefficients are 
*polynomials* in the Jack parameter (`a`): there's no fraction. Actually this 
fact is always true for any Jack polynomial (for any *J*-Jack polynomial, 
I should say). This is an established fact and it is not obvious (it is a 
consequence of the [Knop & Sahi formula][KnopSahi]).

You can substitute a value to the Jack parameter with the help of the 
`substituteParameters` function:

```{r}
( J5 <- substituteParameters(J, 5) )
J5 == JackPol(2, lambda = c(3, 1), alpha = "5")
```

Note that you can change the letters used to denote the variables. By default, 
the Jack parameter is denoted by `a` and the variables are denoted by `X`, `Y`, 
`Z` if there are no more than three variables, otherwise they are denoted by 
`X1`, `X2`, ... Here is how to change these symbols:

```{r}
showSymbolicQsprayOption(J, "a") <- "alpha"
showSymbolicQsprayOption(J, "X") <- "x"
J
```

If you want to have the variables denoted by `x` and `y`, do:

```{r}
showSymbolicQsprayOption(J, "showMonomial") <- showMonomialXYZ(c("x", "y"))
J
```


## Compact expression of Jack polynomials

The expression of a Jack polynomial in the canonical basis can be long. Since 
these polynomials are symmetric, one can get a considerably shorter expression 
by writing the polynomial as a linear combination of the monomial symmetric 
polynomials. This is what the function `compactSymmetricQspray` does:

```{r}
( J <- JackPol(3, lambda = c(4, 3, 1), alpha = "2") )
cat(compactSymmetricQspray(J))
```

The function `compactSymmetricQspray` is also applicable to a `symbolicQspray`
object, like a Jack polynomial with symbolic Jack parameter.

It is easy to figure out what is a monomial symmetric polynomial: `M[i, j, k]` 
is the sum of all monomials `x^p.y^q.z^r` where `(p, q, r)` is a permutation 
of `(i, j, k)`.


<!-- -------------------- links -------------------- -->

[JackPolynomials]: https://en.wikipedia.org/wiki/Jack_function "Jack polynomials on Wikipedia"

[qspray]: https://github.com/stla/qspray "the 'qspray' package on Github"

[symbolicQspray]: https://github.com/stla/symbolicQspray "the 'symbolicQspray' package on Github"

[KnopSahi]: https://en.wikipedia.org/wiki/Jack_function#Combinatorial_formula "Knop & Sahi formula"
